Experience Report: Haskell in Computational Biology

### 2012-03-05 mrfy_experience_report.tex
further changes to viterbi section. more to come.

%% move into later section

%% add something about groups of related proteins aligned (put into correspondence)

@@

%% how do proteins change due to evolution?
%% explain insertion, deletion, substitution.

%% what looks like an intimidating geometry is completely explained by the fact
%% that we have insertions, deletions, and substitutions.

>>>>

In typical use, HMMER builds its structural template -- its Markov random field 
-- from a group of proteins that share common function and structure. A 
biologist puts these protein sequences into a correspondence, or alignment, 
typically using structural alignment software. The columns of the alignment 
(Figure~\ref{alignment}) may contain gaps, indicating that a particular protein 
does not have an amino acid, or residue, participating in that position. Gaps 
in a few sequences at a position likely correspond to evolutionary deletions, 
while gaps in many sequences at a position may correspond to evolutionary 
insertions. A position that has residues in a majority of the sequences becomes 
a \textit{match} state in the hidden Markov model; this position may see a 
variety of amino acids due to evolutionary substitution, but it still 
corresponds to a position that a homologous protein should most likely fill.

The resulting match states, as well as \textit{insertion} states that allow 
additional residues to appear in a protein sequence, and \textit{deletion} 
states that allow match states to be skipped, form a directed graph, the hidden 
Markov model (Figure~\ref{plan7}). Each match or insertion state has an 
associated table of emission probabilities, which correspond to the probability 
of seeing each specific amino acid letter at a given position. Triplets of 
insertion, deletion, and match states are grouped into ``nodes,'' which 
correspond to columns of the original alignment. What looks like an 
intimidating geometry is thus completely explained by the fact that we have 
insertions, deletions, and substitutions due to evolution.

%% now explain computing a *parse* (HMMER) leading into Viterbi.

@@
